---
title: "Project 2"
author: "Team Members: Adrian Missier (USPRT) & Kiyoko Shimizu (WHO) "
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include = FALSE}

#### Load necessary packages ####
options(repos = c(CRAN = "https://cloud.r-project.org/"))
if(!require(tidyverse)){install.packages("tidyverse")}
if(!require(knitr)){install.packages("knitr")}
if(!require(kableExtra)){install.packages("kableExtra")}
if(!require(leaflet)){install.packages("leaflet")}
if(!require(DT)){install.packages("DT")}
if(!require(cowplot)){install.packages("cowplot")}

library(tidyverse)
library(knitr)
library(kableExtra)
library(leaflet)
library(DT)
library(cowplot)
```

## Background
The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases. 
*CSIT-165* disseminates data driven analyses to global decision makers.

*CSIT-165* is a conglomerate comprised of two fabricated entities: World Health Organization (WHO) and U.S. Pandemic Response Team (USPRT). 
We have been assigned the role of data scientist for the entities as follows:
Adrian Missier - USPRT;
Kiyoko Shimizu - WHO

## Data
> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)

Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE). Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. 
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user. Date fields are stored in MM/DD/YYYY format.

```{r WHO_datasets_download}
# URLs for global data sets for COVID-19 associated confirmations and deaths 
deaths_global_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
  
confirmed_global_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"

# Create curl command
deaths_global_command<-paste("curl",deaths_global_url, "-o", "time_series_covid19_deaths_global.csv")
global_command <- paste("curl", confirmed_global_url, "-o", "time_series_covid19_confirmed_global.csv")

# Run the commands
system(deaths_global_command)
system(global_command)

# Read data from .csv files
deaths_global_data <- read.csv("time_series_covid19_deaths_global.csv", stringsAsFactors = FALSE)
confirmed_global_data <- read.csv("time_series_covid19_confirmed_global.csv", stringsAsFactors= FALSE)  
```
```{r USPRT_datasets_download}
# URLs for US data sets for COVID-19 associated confirmations and deaths 
confirmed_us_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"

deaths_us_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"

# Create curl command
deaths_us_command <- paste("curl", deaths_us_url, "-o", "time_series_covid19_deaths_US.csv")
us_command <- paste("curl", confirmed_us_url, "-o", "time_series_covid19_confirmed_US.csv")

# Run the commands
system(deaths_us_command)
system(us_command)

# Read data from .csv files
confirmed_us_data <- read.csv("time_series_covid19_confirmed_US.csv", stringsAsFactors= FALSE)
deaths_us_data <- read.csv("time_series_covid19_deaths_US.csv", stringsAsFactors = FALSE)
```
```{r prepare_data}
# Clean and prepare global data
clean_global_data <- function(data) {
  cleaned <- data %>%
    filter(!is.na(Lat), !is.na(Long), Lat != 0, Long != 0) 
  return(cleaned)
}

confirmed_global_cleaned <- clean_global_data(confirmed_global_data)
deaths_global_cleaned <- clean_global_data(deaths_global_data)

# Filter out cruise ships and invalid coordinates based on Lat/Long
confirmed_global_cleaned <- confirmed_global_cleaned %>%
  filter(!is.na(Lat), !is.na(Long), Lat != 0, Long != 0)
  #filter(if_all(where(is.numeric), ~ . >= 0))

deaths_global_cleaned <- deaths_global_cleaned %>%
  filter(!is.na(Lat), !is.na(Long), Lat != 0, Long != 0)
  #filter(if_all(where(is.numeric), ~ . >= 0))
```
## Project Objectives

### Objective 1: Global Map - Covid-19 Confirmations & Deaths
```{r ob1}
# Identify the column with last date
date_last <- names(confirmed_global_cleaned)[ncol(confirmed_global_cleaned)]

# Sum confirmations and deaths of provinces into one value for each country they belong to 
confirmed_global_summed <- confirmed_global_cleaned %>% 
  group_by(Country.Region) %>% 
  summarise(confirmed = sum(.data[[date_last]], na.rm = TRUE),
            Lat = mean(Lat, na.rm = TRUE),
            Long = mean(Long, na.rm = TRUE))
            
deaths_global_summed <- deaths_global_cleaned %>% 
  group_by(Country.Region) %>% 
  summarise(deaths = sum(.data[[date_last]], na.rm = TRUE))

# Join confirmed and deaths data frames
global_joined <- left_join(confirmed_global_summed, deaths_global_summed, by = "Country.Region")

# Find the first, second and third quartiles
quartiles_confirmed <- quantile(global_joined$confirmed, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
quartiles_deaths <- quantile(global_joined$deaths, probs = c(0.25,0.5,0.75), na.rm = TRUE)

# Set colors for quartiles
global_joined  <- global_joined %>% 
  mutate(
    conf_color = case_when(
    confirmed <= quartiles_confirmed[1] ~ "blue",
    confirmed <= quartiles_confirmed[2] ~"gray",
    confirmed <= quartiles_confirmed[3] ~ "red",
    ),
    deaths_color = case_when(
      deaths <= quartiles_deaths[1] ~ "blue",
      deaths <= quartiles_deaths[2] ~"gray",
      deaths <= quartiles_deaths[3] ~ "red",
    )
  )
# Make map using leaflet
leaflet(global_joined) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  
  # Add layer for confirmed cases
  addCircleMarkers(
    data = global_joined,
    group = "confirmed",
    lng =~ Long,
    lat = ~Lat,
    radius = ~log1p(confirmed)/2,
    color = ~conf_color,
    label = ~`Country.Region`,
    popup = ~paste0("<b>Country:</b> ", `Country.Region`, "<br>",
                   "<br><b>Confirmations:</b> ", confirmed,
                   "<br><b>Deaths:</b> ",deaths),
    stroke = FALSE, fillOpacity = 0.6
  ) %>% 
  
  # Add layer for deaths
  addCircleMarkers(
    data = global_joined,
    group = "deaths",
    lng = ~Long, lat = ~Lat,
    radius = ~log1p(deaths)/2,
    color = ~deaths_color,
    label = ~`Country.Region`,
    popup = ~paste0("<b>Country:</b> ", `Country.Region`, "<br>",
                   "<b>Deaths:</b> ", deaths),
    stroke = FALSE,fillOpacity = 0.6
  ) %>%
  
  # Add checkboxes
  addLayersControl(
    overlayGroups = c("confirmed","deaths"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% 
  hideGroup("deaths")
```
### Objective 2: Table with Top 25 Countries with Highest Number of  Confirmed Cases and Deaths
```{r ob2}
# Identify top 25 countries with highest number of confirmed cases
confirmed_top25 <- global_joined %>% 
  arrange(desc(confirmed)) %>% 
  slice(1:25) %>% 
  select(Country = `Country.Region`, Count = confirmed)

# Identify top 25 countries with highest number of deaths
deaths_top25 <- global_joined %>% 
  arrange(desc(deaths)) %>% 
  slice(1:25) %>% 
  select(Country = `Country.Region`, Count = deaths)

# Combine confirmed_top and deaths_top tables and add new column for rank number
table_ranked <- tibble(
  Rank = 1:25,
  country_confirmed = confirmed_top25$Country,
  country_count = confirmed_top25$Count,
  country_deaths = deaths_top25$Country,
  count_deaths = deaths_top25$Count
)

# Rename columns for table_ranked table
colnames(table_ranked) <- c("Rank", "Country", "Count", "Country", "Count")

# Display new table 
table_ranked %>% 
  kable(caption = "Top 25 Countries by Confirmed Cases and Deaths",
        col.names = c("Rank", "Country", "Count", "Country", "Count")) %>% 
  kableExtra::add_header_above(c(" " = 1, "Confirmations" = 2, "Deaths" = 2)) %>% 
  kableExtra::kable_styling(bootstrap_options = c("condensed"))
```
### Objective 3: Covid-19 Confirmations in California
```{r ob3}
library(lubridate)
library(cowplot)

# Filter to get California
ca_confirmed <- confirmed_us_data %>%
  filter(Province_State == "California")

latest_date <- names(ca_confirmed)[ncol(ca_confirmed)]

# Get top 3 counties in California with the most confirmed cases
top_3_counties <- ca_confirmed %>%
  select(Admin2, all_of(latest_date)) %>%
  arrange(desc(.data[[latest_date]])) %>%
  slice_head(n = 3) %>%
  pull(Admin2)

# Build a Confirmed Case Trend Line in California
ca_total <- ca_confirmed %>%
  select(matches("^X\\d")) %>%
  summarise(across(everything(), sum)) %>%
  pivot_longer(cols = everything(), names_to = "Date", values_to = "Cases") %>%
  mutate(Date = mdy(str_remove(Date, "^X")))

# Convert the top 3 counties into long form data
top_cities <- ca_confirmed %>%
  filter(Admin2 %in% top_3_counties) %>%
  select(Admin2, matches("^X\\d")) %>%
  pivot_longer(cols = matches("^X\\d"), names_to = "Date", values_to = "Cases") %>%
  mutate(Date = mdy(str_remove(Date, "^X")))

# Important dates on timeline
milestones <- tibble(
  label = c("SAH Order", "EUA", "+6 weeks", "Delta", "Omicron"),
  date = mdy(c("03-19-2020", "01-29-2021", "03-12-2021", "05-11-2021", "11-26-2021"))
)

# 1st graph with total confirmations for California
plot_ca <- ggplot(ca_total, aes(x = Date, y = Cases)) +
  geom_line(color = "darkblue", linewidth = 1) +
  labs(
    title = "COVID-19 Confirmations in California",
    x = "Date", y = "Confirmations"
  ) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b\n%Y") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5)) +
  geom_vline(data = milestones,
             aes(xintercept = date, color = label),
             linetype = "dashed", linewidth = 1) +
  geom_text(
    data = milestones,
    aes(x = date, y = max(ca_total$Cases) * 1.05, label = label),
    size = 2.5,
    fontface = "bold",
    nudge_x = -15
  ) +
  scale_color_manual(values = c(
    "SAH Order" = "darkred",
    "EUA" = "blue",
    "+6 weeks" = "green",
    "Delta" = "purple",
    "Omicron" = "orange"
  )) +
  guides(color = "none")

# 2nd graph with Top 3 Counties
plot_top_3 <- ggplot(top_cities, aes(x = Date, y = Cases, color = Admin2)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Top 3 California Counties by Confirmed Cases",
    x = "Date",
    y = "Cases",
    color = "County"  # Renames the legend title from 'Admin2' to 'County'
  ) +
  theme_minimal() +
  geom_vline(data = milestones, aes(xintercept = date), linetype = "dashed", color = "red") +
  geom_text(
    data = milestones,
    aes(x = date, label = label),
    y = Inf,
    angle = 90,
    vjust = -0.5,
    hjust = 0,
    inherit.aes = FALSE
  )

plot_grid(plot_ca, plot_top_3, ncol = 1)
```
### Objective 4: Exploring Relationship between -  Population & Confirmed Counts, Death Counts & Confirmed Counts
```{r ob4}
library(dplyr)
library(ggplot2)
library(cowplot)
library(scales)

# Filter California to get Confirmed and Deaths
confirmed_ca <- confirmed_us_data %>%
  filter(Province_State == "California")

deaths_ca <- deaths_us_data %>%
  filter(Province_State == "California")

latest_date <- names(confirmed_ca)[ncol(confirmed_ca)]

# Popualtion for California counties from 2020 Census
ca_pop <- tibble::tribble(
  ~Admin2, ~Population,
  "Los Angeles", 10039107,
  "San Diego", 3338330,
  "Orange", 3175692,
  "Riverside", 2470546,
  "San Bernardino", 2180085,
  "Santa Clara", 1927852,
  "Alameda", 1671329,
  "Sacramento", 1552058,
  "Contra Costa", 1153526,
  "Fresno", 999101,
  "Kern", 900202,
  "San Francisco", 881549,
  "Ventura", 846006,
  "San Mateo", 766573,
  "San Joaquin", 762148,
  "Stanislaus", 550660,
  "Sonoma", 494336,
  "Tulare", 466195,
  "Solano", 447643,
  "Santa Barbara", 446499,
  "Monterey", 434061,
  "Placer", 404739,
  "San Luis Obispo", 283111,
  "Santa Cruz", 273213,
  "Merced", 277680,
  "Marin", 258826,
  "Butte", 219186,
  "Yolo", 220500,
  "El Dorado", 192843,
  "Shasta", 180080,
  "Imperial", 181215,
  "Madera", 157327,
  "Kings", 152940,
  "Napa", 137744,
  "Humboldt", 135558,
  "Nevada", 99755,
  "Sutter", 99424,
  "Mendocino", 86749,
  "Yuba", 78668,
  "Lake", 64386,
  "San Benito", 62808,
  "Tuolumne", 54478,
  "Calaveras", 45905,
  "Siskiyou", 43539,
  "Amador", 39752,
  "Lassen", 30573,
  "Glenn", 28917,
  "Del Norte", 27812,
  "Colusa", 21547,
  "Plumas", 18807,
  "Tehama", 65357,
  "Mariposa", 17203,
  "Inyo", 18039,
  "Trinity", 12285,
  "Mono", 14444,
  "Modoc", 8841,
  "Sierra", 3005,
  "Alpine", 1129
)

# Merge population data with confirmed and deaths
confirmed_ca <- left_join(confirmed_ca, ca_pop, by = "Admin2")
deaths_ca <- left_join(deaths_ca, ca_pop, by = "Admin2")

# Combine the data so there is one dataset for California containing deaths, confirmations, population
covid_summary <- confirmed_ca %>%
  select(Admin2, Population, !!latest_date) %>%
  rename(Confirmed = !!latest_date) %>%
  left_join(
    deaths_ca %>%
      select(Admin2, !!latest_date) %>%
      rename(Deaths = !!latest_date),
    by = "Admin2"
  ) %>%
  filter(Population > 0, Confirmed > 0, Deaths > 0) %>%
  mutate(
    log_Population = log10(Population),
    log_Confirmed = log10(Confirmed),
    log_Deaths = log10(Deaths)
  )

# Plot 1 - Confirmations vs Population
plot1 <- ggplot(covid_summary, aes(x = log_Population, y = log_Confirmed)) +
  geom_point(color = "steelblue") +
  labs(
    title = "Confirmations vs Population",
    x = "Population",
    y = "Confirmed Cases"
  ) +
  scale_x_continuous(labels = function(x) scales::comma(10^x)) +
  scale_y_continuous(labels = function(x) scales::comma(10^x)) +
  theme_gray(base_size = 13)

# Plot 2 - Deaths vs Confirmations
plot2 <- ggplot(covid_summary, aes(x = log_Confirmed, y = log_Deaths)) +
  geom_point(color = "firebrick") +
  labs(
    title = "Deaths vs Confirmed Cases",
    x = "Confirmed Cases",
    y = "Deaths"
  ) +
  scale_x_continuous(labels = function(x) scales::comma(10^x)) +
  scale_y_continuous(labels = function(x) scales::comma(10^x)) +
  theme_gray(base_size = 13)

# Display both plots together using cowplot
plot_grid(plot1, plot2, ncol = 2)
```

### GitHub Log
```{bash gitlog } 
git log --pretty=format:"%nSubject: %s%nAuthor: %aN%nDate: %aD%nBody: %b"
```
[GitHub repository] 

